---
title: "vimes: a short demonstration"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{demonsration}
  \usepackage[utf8]{inputenc}
---



```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 6, 
  fig.path = "figs-demo/"
)

```

This vignette provides a short demonstration of the package using a dummy
dataset.

We first simulate the data using 3 mixtures of 3 normal distributions, and
compute Euclidean distances between the observations for each mixture.  In
practice, each mixture would be a different data type (e.g. location, time of
onset of symptoms, genetic sequences of the pathogen):

```{r, data}
set.seed(2)
dat1 <- rnorm(30, c(0,1,6))
dat2 <- rnorm(30, c(0,0,1))
dat3 <- rnorm(30, c(8,1,2))
x <- lapply(list(dat1, dat2, dat3), dist)
```

The function `vimes_data` processes the data and ensures matching of the
individuals across different data sources:

```{r, vimesdata}
library(vimes)

x <- vimes_data(x)
plot(x)

```

We can now run `vimes` on the data:
```{r, vimes}

res <- vimes(x, cutoff = c(2,4,2))
names(res)

res$graph
res$clusters
```

The main graph is:
```{r, res}
plot(res$graph, main="Main graph")
for(i in 1:3) {
plot(res$separate_graphs[[i]]$graph, main = paste("Graph from data", i))
}
```