---
title: "Expected distributions of pairwise distances."
author: "Thibaut Jombart, Anne Cori"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{distributions}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 5, 
  fig.path = "figs-dist/"
)

```


*vimes* uses reference distributions of (spatial, temporal, genetic) distances
 between pairs of cases on a transmission chain to define thresholds for the
 graph pruning algorithm. In such case, we asssume these distributions are
 known, typically from the literature. However, these distributions change when
 there are unobserved cases. In this vignette, we illustrate how the
 distribution of expected distances between cases is computed for a give
 reporting probability `pi`, and given a known distribution *F*(*d*).



Distributions of distances with missing cases
---------------------------------------------

Let us assume a simple probability mass function of, say, the *serial
interval*. We simulate it using the package *distcrete*, from a discretised
Gamma distribution with shape 10 and rate 1.5:

```{r, serial}

si <- distcrete::distcrete("gamma", 1L,
                           shape = 10,
			   rate = 1.2)$d(1:20)

plot(si, type = "h", lwd = 10, lend=2,
     xlab = "Days after onset",
     main = "Serial interval distribution")

```

This distribution assumes a single generation of infection between cases. Its
general form for a number *K* of generations between cases is given by $K$
convolutions of *F*(*d*). In *vimes*, these are internally computed by
`convolve_empirical`. As the function is not exported, one needs to use:

```{r, convolve}

si_10 <- vimes:::convolve_empirical(si, 10, TRUE)

```

to obtain *K*=10 convolutions of the distribution (`si`); the last argument is
to retain all intermediate convolutions for *k*=1, ..., K. 

```{r, convolve2}

head(si_10)
apply(si_10, 2, sum)

matplot(si_10[1:120,], type = "l", lty = 1, lwd = 2, col = rainbow(10),
	xlab = "Days after onset",
	main = "Serial interval, k = 1, ..., 10 missing cases")
mtext("from k = 1 (red) to k = 10 (purple)", 3)

```






Integrating over missing cases
------------------------------

In practice, we do not know the number of generations between two cases, so we
cannot use any of the convolutions above directly. However, it is possible to
integrate over all possible values of *K*. In practice, we will limit the
convolutions to a maximum value of *K* so that *p*(*k* > *K*) < threshold,
defaulting to 0.001. The resulting distribution is a linear combination of the
convolutions, weighted by the likelihoods of the corresponding values of
*k*. These weights are computed for a given reporting probability `pi` by, for
instance:

```{r, weights}

w <- vimes:::get_weights(0.6, 10)
plot(w, type = "h", xlab = "k", ylab = "weight",
     lwd = 10, lend = 2, main = "Weights")

```

for a reporting probability of 0.6 and a maximum value of *k* of 10.  The final
distribution combines the previous convolutions using these weights. This is
achieved by the (exported) function `dempiric`:

```{r, final}

d_final <- dempiric(si, 0.6)
head(d_final)
sum(d_final)

pal <- colorRampPalette(c("navy", "lightblue", "grey"))

plot(d_final[1:80], type = "h", lwd = 3, lend = 2,
     col = pal(80), xlab = "Days after onset",
     main = "Distributions of temporal distances for pi = 60%")

```