---
title: "vimes: an illustration on a simulated dog rabies dataset"
author: "Anne Cori & Thibaut Jombart"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{demonsration}
  \usepackage[utf8]{inputenc}
---



```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 6, 
  fig.height = 6, 
  fig.path = "figs-demo/"
)

```

This vignette provides a demonstration of the vimes package and more specifically how it can be used to perform analyses such as those presented in XXXXXXXX [CITE THE PAPER]. Here we use a simulated dataset of rabies transmitted among dogs, chosen among all the baseline simulations described in XXXXXXXX [CITE THE PAPER] as the only simulated dataset with 151 observed dogs infected with rabies, i.e. exactly the same number as in the real dataset analysed in this paper. 

## Loading and examining the data

The simulated dataset is available as  `sim_rabies` in the vimes package:

```{r}
library(vimes)
data(sim_rabies)
sim_rabies$n

```

We can plot the epidemic curve using the `incidence` package:
```{r}
### if incidence is not already installed, run these few lines:
# install.packages("devtools")
# library(devtools)
# install_github("reconhub/incidence")

### load incidence package and use it to plot epidemic curve over different time scales:
library(incidence) 
# weekly
plot(incidence(sim_rabies$onset,7))
# monthly
plot(incidence(sim_rabies$onset,365/12))
# bimonthly
plot(incidence(sim_rabies$onset,365/6))
```

And we can plot the locations of the cases:
```{r}
locations <- t(sapply(1:sim_rabies$n, function(i) sim_rabies$cases[[i]]$location))
plot(locations, xlab = "", ylab="")
```

XXXXXXXXX THIBAUT CAN WE PLEASE ADD A PHYLOGENETIC TREE FOR THIS?


## Computing pairwise distances between cases in time, space, and genetics 


Distances between dates are computed as numbers of days:
```{r}
head(sim_rabies$onset, 5)
D_dates <- dist(sim_rabies$onset)
```

Distances between locations are computed using the great circle distance:
```{r}
library(fields)
head(locations)
D_geo <- rdist(locations)
```

Distances between sequences are computed as simple Hamming distances using function `dist_dna` in the `quicksim` package:
```{r}
#install_github("thibautjombart/quicksim")
library(quicksim)
dna <- lapply(1:sim_rabies$n, function(i) sim_rabies$cases[[i]]$dna)
# dna lists the mutation points compared to a reference genome for the pathogen sequence in each case
head(dna)
D_dna <- dist_dna(dna)
``` 

We format, match and plot the distance data using \emph{vimes}: XXX WHAT IS THE MATCHING USING HERE AS NOT USING IDs I THINK ?!
```{r} 
D_all <- vimes_data(dates = D_dates, geo = D_geo, dna = D_dna)
plot(D_all, nclass = 60)
``` 


## Defining cutoff distances above which cases are considered not linked by transmission 

### Distributions of expected distances between cases for rabies

We retrieved preexisting information about the serial interval, spatial kernel, and mutation rate for rabies, 
and used this information to define the distribution of expected distances between cases in the temporal, spatial, and genetic spaces. 

The serial interval and spatial kernel distributions are taken from Hampson et al. \url{http://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.1000053}
```{r} 
# serial interval distribution parameters
gamma_mean <- 23.55
gamma_std <- 20.85
# convert into shape and scale 
gamma_shape <- gamma_mean^2 / (gamma_std^2)
gamma_scale <- gamma_std^2 / gamma_mean
```

XXXXXXXX NEED TO EXPLAIN MORE WHAT WE DO JUST BELOW AND MAYBE REMOVE RAYLEIGH_SD AS NOT USED?
```{r} 
# spatial kernel parameters
rayleigh_mean <- 0.88
# find Rayleigh parameters to match this mean
rayleigh_scale <- rayleigh_mean / sqrt(acos(-1)/2)  # using acos(-1) instead of pi as we later use pi for reporting rate
rayleigh_sd <- sqrt( rayleigh_scale^2*(4-acos(-1))/2 )  # using acos(-1) instead of pi as we later use pi for reporting rate
```

The mutation rate is derived from Bourhy et al. \url{http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1005525}
```{r} 
# mutation rate
mu_year_per_site <- 5.9e-4
n_sites <- 5061
# mutation rate per day and sequence
mu_day_whole <- (mu_year_per_site * n_sites / 365) 
```

We use the `fpaircase` function to compute the distributions of expected distances (temporal, spatial and genetic respectively) between a case and their infector based on the above parameters. 
```{r} 
## distance functions for each of the 3 types
f_temporal <- fpaircase(type = "temporal", gamma_shape = gamma_shape,
                        gamma_scale = gamma_scale)

f_spatial <- fpaircase(type = "spatial", sd_spatial = rayleigh_scale)

f_genetic <- fpaircase(type = "genetic", poisson_rate = mu_day_whole,
                       gamma_shape = gamma_shape, 
                       gamma_scale = gamma_scale)

# Plotting these
par(mfrow=c(3,1))
plot(f_temporal, xlim = c(0,365))
plot(f_spatial, xlim = c(0,5))
plot(f_genetic, xlim = c(0,5))
```

### Cutoff distances defined as the quantiles of the distributions of expected pairwise distances

We will be using the above distributions of expected distances between a case and their infector to define cutoff distances above which pairs of cases are considered not linked by transmission. 

First, we assume a certain level of reporting. Here we assume only $20\%$ of cases are observed. 
```{r} 
# reporting rate
pi <- 0.2
```

Then, we define the quantile we want to use to define the cutoff distances. Here, we consider many different quantiles to assess sensitivity of our results to this choice. 
```{r} 
# quantiles
q <- c(.50, .75, .90, .95, .95^(1/3), .99, .999, .9995, .9999)
# colours used to plot these
cols <- rainbow(length(q))
cutoff_choice <- q[4] # our main results are with the cutoff corresponding to the 95% quantile
```

We now plot the distributions of expected distances (temporal, spatial and genetic respectively) between a case and their closest observed ancestor, assuming that only $20\%$ of cases are reported. We also show the cutoffs corresponding to the quantiles defined above. 
```{r} 
## distance functions for each of the 3 types, accounting for reporting probability pi
## with quantiles q overlayed on the graphs
par(mfrow=c(3,1))
plot(f_temporal, q, xlim = c(0,365*2), pi = pi,
     lines_arg = list(col=cols, lwd=2))
plot(f_spatial, q, xlim = c(0,10), pi = pi,
     lines_arg = list(col=cols, lwd=2))
plot(f_genetic, q, xlim = c(0,10), pi = pi,
     lines_arg = list(col=cols, lwd=2))
```

We can overlay the graph above on top of the histogram of observed distances to see where these cutoffs fall with respect to our observations. 
```{r} 

### function used to generate a plot with distribution of observed and expected distances
plot_overlay <- function(dist, f, q, pi, xlab, breaks, resol = 1,
                         q_color = cols, hist_bordercolor = "grey", hist_color = "lightgrey")
{ # dist contains the observed distances
  # f is the distribution of expected distances between a case and their infector
  # q is the quantile or vector of quantiles of interest
  # pi is the reporting rate
  # xlab is the x axis label
  # breaks is the breaks used for plotting the histogram of dist
  # resol is the resolution used for plotting f
  # q_color is the colour or vector of colours used to show the cutofs associated with the quantile(s) q
  # hist_bordercolor is the colour used for the border of the histogram of dist
  # hist_color is the colour used for the filling of the histogram of dist
  qtl <- get_quantiles(f, q, pi = pi)
  hist(dist, col = hist_color, border = hist_bordercolor, 
       main="", xlab=xlab, 
       breaks=breaks)
  par(new = T)
  x <- seq(min(breaks), max(breaks), resol)
  y <- f(x, pi = pi)
  plot(x, y, type="l", axes=FALSE, main = "", xlab="", ylab="")
  # add vertical lines corresponding to quantiles
  abline(v = qtl, col = q_color, lwd = 2)
}

### use the function above to create our plot: 
par(mfrow=c(3, 1), mar=c(5, 5, 0.5, 5))

# temporal
plot_overlay(dist = as.vector(D_all$dates), 
             f = f_temporal, 
             q = q, 
             pi = pi, 
             xlab = "Pairwise distance in time (days)", 
             breaks = seq(0,2500, 50),
             resol = 1)

# spatial
plot_overlay(dist = as.vector(D_all$geo), 
             f = f_spatial, 
             q = q, 
             pi = pi, 
             xlab = "Pairwise distance in space (km)", 
             breaks = seq(0,20,0.2),
             resol = 0.1)

# genetic
plot_overlay(dist = as.vector(D_all$dna), 
             f = f_genetic, 
             q = q, 
             pi = pi, 
             xlab = "Pairwise distance in space (km)", 
             breaks = seq(0,55,1),
             resol = 1)
```

## Running vimes to identify clusters of cases linked by transmission

We run vimes to identify clusters of cases linked by transmission, for various cutoff choices, and various reporting rates. We plot the results by colouring cases identified as belonging to the same outbreak cluster, i.e. cases identified as being linked by local transmission. Cases shown in grey are identified as singletons who are not linked by transmission to any other observed case. 
```{r} 
### function used to get results for a certain cutoff and reporting rate
get_res <- function(D_all, q, pi, f_temporal, f_spatial, f_genetic, type=c("all", "temporal","spatial", "genetic")) {
  
  type <- match.arg(type)
  
  ## get the cutoffs
  cuts <- c(temporal = get_quantiles(f_temporal, q, pi = pi), 
            spatial = get_quantiles(f_spatial, q, pi = pi),
            genetic = get_quantiles(f_genetic, q, pi = pi))
  
  if(type == "all")
  {
    ## use vimes
    out <- vimes(D_all, cutoff = cuts,
          graph.opt = vimes.graph.opt(col.pal=funky))
  }
  else if(type == "temporal")
  {
    out <- vimes(vimes_data(dates = D_all$dates), cutoff = cuts$temporal,
                 graph.opt = vimes.graph.opt(col.pal=funky))
  }else if(type == "spatial")
  {
    out <- vimes(vimes_data(geo = D_all$geo), cutoff = cuts$spatial,
                 graph.opt = vimes.graph.opt(col.pal=funky))
  }else if(type == "genetic")
  {
    out <- vimes(vimes_data(dna = D_all$dna), cutoff = cuts$genetic,
                 graph.opt = vimes.graph.opt(col.pal=funky))
  }
    
  return(out)
  
}

### use the function above to generate results for several combinations of p and pi
### We assume, as in Bourhy et al., a reporting rate of 20% in our main analyses, and reporting rates of 10 and 50% respectively in two extreme scenarios considered in sensitivity analyses. 

combi <- expand.grid(p = q,
                     pi = pi)
combi

quantile_pretty <- signif(combi$p*100, 4)
quantile_pretty <- paste0(quantile_pretty, "%")

res <- vector(7L, mode = "list")

for (i in 1:nrow(combi)) {
  res[[i]] <- get_res(D_all, combi[i, 1], combi[i, 2], f_temporal, f_spatial, f_genetic)
}

### visualise the output
par(mfrow = c(3, 3), mar=c(1,1,3,1))

for(i in 1:length(res)) {
  plot(res[[i]]$graph, vertex.label="",
       main = paste(
         "cutoff:", quantile_pretty[i]))
}

```

## Compare vimes results to 'truth'

Since we are using simulated data, we can actually compare the results obtained using `vimes` to the truth, i.e. check whether cases identified as belonging to the same outbreak cluster were indeed part of the same transmission chain in the simulation

```{r} 
i <- which(combi$p %in% 0.95) # taking this is our main result
# cluster membership as identified by vimes
est_clust_membership <- as.vector(res[[i]]$clusters$membership)
# cluster membership as simulated
sim_clust_membership <- sim_rabies$clust_renumbered

est_clust_membership
sim_clust_membership
```

For a formal comparison, we can measure the proportion of pairs of individuals belonging to the same outbreak cluster which were indeed clustered together by the method (True Positive Rate), as well as the proportion of pairs of individuals which did not belong to the same outbreak cluster and were adequately placed in different groups by the method (True Negative Rate).
```{r} 
# function to compute the true positive and true negative rates
get_TPR_TNR <- function(est_clust_membership, sim_clust_membership)
{

### look at pairs of individuals and whether they are in the same cluster or not in simulation and estimation
  sim_same_clust <- as.matrix(dist(sim_clust_membership, method="manhattan"))==0
  est_same_clust <- as.matrix(dist(est_clust_membership, method="manhattan"))==0
  
  sim_same_clust[lower.tri(sim_same_clust, diag = TRUE)] <- NA
  est_same_clust[lower.tri(est_same_clust, diag = TRUE)] <- NA
  
  sim_same_clust_vect <- na.omit(as.vector(sim_same_clust))
  est_same_clust_vect <- na.omit(as.vector(est_same_clust))

### see https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12968 for the following param definitions
  ## The true positive rate (TPR) 
  # defined as the proportion of individuals belonging to the same population which were indeed clustered together by the method.
  TPR <- sum(est_same_clust_vect[which(sim_same_clust_vect)]) / sum(sim_same_clust_vect)
  
  ## The true negative rate (TNR) 
  # defined as the proportion of individuals which did not belong to the same population and were adequately placed in different groups by the method
  TNR <- sum(!est_same_clust_vect[which(!sim_same_clust_vect)]) / sum(!sim_same_clust_vect)
  
  return(c(TPR = TPR, TNR = TNR))
}

# use this function to compute the TPR and TNR for all the cutoffs we considered
all_TPR_TNR <- as.data.frame(t(sapply(1:length(res), function(i) get_TPR_TNR( as.vector(res[[i]]$clusters$membership), sim_clust_membership) )))

all_TPR_TNR

all_TPR_TNR$quantile <- combi$p
all_TPR_TNR$SumTPRTNR <- all_TPR_TNR$TPR + all_TPR_TNR$TNR

# plot results as a ROC curve
plot(1 - all_TPR_TNR$TNR, all_TPR_TNR$TPR, xlab = "1 - TNR", ylab = "TPR", xlim = c(0, 1), ylim = c(0, 1), col = cols, pch = 19, cex = 1.5)
legend("bottomright", legend = quantile_pretty, title = "Quantile used for cutoff", col = cols, pch = 19, pt.cex = 1.5, bty = "n", cex = 0.75)
```

## Estimate the underlying reproduction number and number of imported cases

XXXXXX TO DO

## Other things to add

- Maybe show more specifically how wrong we are when using quantile 0.95 rather than optimum 0.99%? (could be shown by producing Figure similar to Figure 2 for both sets of results; see next point)

- Show the clusters on graphs in time, space and genetics as in paper (Figure 2)

- Show results obtained with one data stream only to illustrate value of combining data streams (Figure 4 in paper)

- Show real time analyses (Figure S8 in paper)

- Maybe - show results obtained with misspecified reporting rates?






