---
title: "Detecting disease outbreaks using vimes"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{vimes: VIsualisation and Monitoring of EpidemicS.}
  \usepackage[utf8]{inputenc}
---


```{r setup, echo=FALSE}
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=10, fig.path="figs/")
options(digits = 4)
```
*vimes*: VIsualisation and Monitoring of EpidemicS 

=================================================
*vimes* provides tools for integrating various types of surveillance data for detecting disease outbreaks. This document provides an overview of the package's content.


Installing *vimes*
-------------
To install the development version from github:
```{r install, eval=FALSE}
library(devtools)
install_github("thibautjombart/vimes")
```

The stable version can be installed from CRAN using:
```{r install2, eval=FALSE}
install.packages("vimes")
```

Then, to load the package, use:
```{r load}
library("vimes")
```

